Will a~human always outsmart a~computer? An essay

### 2023-02-06 ESS_Olszewski/Olszewski.tex
Final commit before sending to the editors for corrections

\begin{artengenv}{Adam Olszewski}
	{Will a~human always outsmart a~computer? An essay}
	{Will a~human always outsmart a~computer? An essay}
	{Will a~human always outsmart a~computer? An essay}
	{Center for Advanced Internet Studies}
	{The title question of the paper has its empirical origin in the form of an individual's existential experience arising from the personal use of a~computer, which we attempt to describe in the first section. The rest of the entire paper can be understood as a~philosophical essay answering the question posed. First the connection between the main problem of the article and its ``premonition'' by mankind, which was expressed in the form of ancient myths and legends, is briefly suggested. After shortly discussing the problems that early considerations of AI focused on, i.e. whether machines can think at all, we move on to reformulate our title question, about the possibility of outsmarting AI. This outsmarting will be understood by us in a~rather limited way as to prevent a~machine from completing its implemented task. To achieve this objective, after softly clarifying the basic terms, an analogy is built between the ``outsmarting'' of a~machine by a~human (the target domain) and the playing of a~mathematical game between two players (the base domain), where this outsmarting is assigned a~``winning strategy'' in the certain game. This mathematical model is formed by games similar to Banach-Mazur games. The strict theorems of such games are then proved and applied to the target of the analogy. We then draw conclusions and look for counter-examples to our findings. The answer to the title question posed is negative, and it is not clear how far it should be taken seriously.
	}
	{existential experience, myth, computer, machine, Banach-Mazur games, winning strategy.}

@@

\section{An existential ambient of the problem}
The appearance of personal computers—which took place in the middle of 1970s in the world (USA) and a~decade later in Poland (due to the ``iron curtain'' which at that time separated this country from the rest of the world)—brought hope for providing man with a~useful tool with versatile applications and capable of performing certain very practical functions. Thus, in those days, from the perspective of an average user, a~computer standing on his desk was a~\textit{machine}, that is a~concrete, experiential and \textit{human-friendly} device, supporting his activities, for example as a~typewriter or a~memory bank for storing data. Since then, over nearly fifty years, such a~notion of a~useful machine has considerably evolved. This has happened thanks to the extremely dynamic development of information technology (which included increasing the computational capabilities of machines, their speed and memory and developing new and better programming techniques and languages) and the emergence of the Internet. It seems that the present concept of a~\textit{machine} has undergone a~substantial modification. This machine, which was a~friendly tool supporting man and operating in virtual (non-real) time and separated from the outside world, suddenly became a~machine operating in real time and additionally started drawing us deeper and deeper into a~kind of addiction and making us undertake certain activities, which—although performed through a~machine standing on the desk—have far-reaching consequences in real time and, more importantly, in reality. This new machine on our desk is, in fact, a~terminal of an extremely vast network (\textit{inter-net}), in which powerful algorithms create the elements of the \textit{machine proper}. This \textit{new machine}, which has taken on \textit{a~new meaning}, is also interactive in a~double sense. First, it is able to learn in a~certain sense strictly defined by computability theory, and second, it is known that it must be managed by or is in the hands of a~person or a~group of people because it is not an independent entity (or at least we do not know about it yet). We can call such a~machine \textit{artificial intelligence} (AI). We deliberately omit here a~wide array of issues that need to be clarified or systematically listed\footnote{For some clarifications see
%\label{ref:RNDasnA14J7h1}(Krzanowski and Polak, 2022).
\parencite[][]{krzanowski_meta-ontology_2022}.%
}; these shortcomings justify using the term \textit{essay} in the title of this work. Besides, this paper is my first attempt at analyzing this somewhat strange subject, so I~consider it to be the first of a~series of papers and an essayistic outline of my further work on the issue.

@@

\begin{itemize}
\item \textit{ante rem}—in the world of Platonic ideas—an algorithm;
\item \textit{post-ante rem}—as a~construction of the mind—a program;\footnote{\textit{Post-ante} is a~strange term which is to indicate that from one point of view a~program appears before the \textit{in re} phase and after the \textit{ante rem} phase, while from another point of view it appears after the \textit{in re} phase.}
\item \textit{in re}—physically present in the thing—implemented into a~machine (computer) in the form of the processes that control the behavior of the physical device.\footnote{Implementation is an extremely interesting concept in the philosophy of computer science. It allows us to transform something abstract into something physical, which requires thorough consideration. There is a~certain similarity here with the criticism of Platonism, where one of the objections against Platonism is the impossibility to move from the abstract to the physical.}
\end{itemize}
In other words, the three objects under consideration are the same object which manifests itself in three different forms or phases. We are not necessarily going to defend the above distinction at all costs, but we believe that it is easy to understand and will facilitate further reasoning although it will not be essential for the core of our considerations.

@@

\section{Myths as 'premonitions' of mankind}
The existential situation outlined above may—and sometimes does—cause anxiety in some members of the community of human users, as is amply demonstrated in literature. However, it is difficult to point to any facts to which we have access in the form of empirical verification, on the basis of which a~rigorous narrative could be constructed regarding the justification of existential anxiety. On the argumentative side, it is difficult to find such premises that would make it possible to justify a~conclusion regarding a~future threat from AI. Therefore, literary works devoted to this issue are based on speculations and some even treat it as science fiction, while others treat it as a~purely logical possibility (cf. article motto). We, however, do not downplay this premonition of humanity (expressed in literary terms), and in this section we will try to explain where this anxiety may come from. Our explanation refers to myths that appeared in human history a~long time ago. Mankind, as a~species, through its representatives, has created strange stories called \textit{myths}. I~call these stories strange because, being created at a~very early stage in the development of our species, they speak of problems that have been continuously accompanied it in its history. For example, we have the myth of the Sphinx, a~hybrid winged creature with the body of a~lion and the head of a~woman, who killed travelers heading to Thebes if they failed to solve a~riddle. A~similar pattern can be seen in the case of the dragon of Wawel Castle from the legend of Krakus and of many other creatures from various legends. The motif shared by many of these stories is an attempt to defeat or outwit an evil creature, which poses a~threat to the normal functioning of the community in which it appears. The origin of this creature is in some sense beyond the natural. Most often, in these stories an attempt to outwit it is a~rational act, such as solving the riddle of the Sphinx or giving a~dragon a~fake sheep filled with sulfur. Some believe—and these are not only so-called \textit{ordinary people} but also distinguished scientists—that in certain situations AI is \textit{per analogiam} appears as the embodiment of a~\textit{mythical creature}, which threatens the normal functioning of a~society, violating the freedom or privacy of all or some of its members. And, consequently, there is the question of outwitting the creature. Let us repeat that these myths, being a~kind of common heritage of mankind, are the real cause of humans' fears and anxieties.

\section{Turing's and Searle's tests—AI's first issues}
In 1936, the groundbreaking year in the creation of AI,\footnote{Here we are making a~mental shortcut, because strictly speaking it was crucial for AI to create a~computer as a~machine that practically realises the mathematical idea of computability.} two formal models of computability of effectively computable functions were published: one in the work by the American mathematician Alonzo Church (lambda calculus) and the other in the work by the American logician, born in Augustów (Poland) Emil Post (machine). A~year later the most famous work in the area was published: the one written by the English mathematician and logician, Alan Turing (machine). Very soon Turing's theoretical machines found implementation in the form of real machines, which were called \textit{computers}. This started a~discussion about the capabilities of such machines. Turing
%\label{ref:RNDhKwggb6xjl}(1950)
\parencite*[][]{turing_computing_1950} %
 proposed what is known today as \textit{the Turing test}, in which a~human evaluator judges whether his interlocutor in a~conversation held in a~natural language is a~human or a~machine (AI). Turing's intention in this test was to try to answer the question whether a~machine can think like a~human. He considered the theological argument (one of the arguments against machine thinking), according to which God created man as the only thinking being in the universe, and thinking was a~function of the human soul. Without going into the intricacies of the problem here, let us note that according to Catholic theology, thinking belongs to God and to His angels, thus it is not a~function of the body or brain as I~think Turing probably believed.\footnote{From a~certain point of view, calculation does not intrinsically belong to thinking, because, for example, according to Catholic theology, God, although he thinks, does not calculate, because calculation is the manifestation of a~certain kind of ignorance. Cf. for example Isaiah 55:8-9: ``For my thoughts are not your thoughts, neither are your ways my ways'', declares the Lord. ``As the heavens are higher than the earth, so are my [\ldots] thoughts [higher] than your thoughts.'' Some people believe (basing this belief on the Bible) that God's words in which He speaks of cause and effect justify attributing thinking to Him.} Turing wanted to convince his contemporaries that a~machine can think like a~human being in order to contradict underestimation of a~machine's capabilities in this regard, widespread at that time. The second test—called the Chinese Room argument—comes from John Searle 
%\label{ref:RND4igiqQ1wVO}(1980),
\parencite*[][]{searle_minds_1980}, %
 and was intended to demonstrate that no digital computer has a~kind of ``mind'' or ``consciousness'' even if it functionally bears a~far-reaching resemblance to human conversational behaviour. Searle's argument dampened the enthusiasm of optimistic proponents of AI and its unlimited possibilities. It can be said that the texts mentioned in this section share a~common goal: to take a~stand in the argument about machine thinking at a~time when, in popular understanding, a~machine's skills were not appreciated. From the point of view of this paper, both texts (Turing's and Searle's) have lost a~great deal of relevance since their publication, as so much has changed in this area. Summing up what has been said so far briefly and succinctly: today no one asks whether machines can think but rather what machines can do in terms of thinking and intelligence and where the upper limits of their capabilities lie.

\section{Preliminary formulation of the fundamental question}
Taking into account what has been outlined above, we can say that, from a~particular point of view, AI can appear as an element of reality—as an artefact—which poses a~threat to society on the way to the unrestricted realization of its development. Therefore, we should look for means by which we could ‘outsmart' this \textit{creature}. In different words, we can ask if a~human can stand up to AI that controls him.\footnote{An extreme case of this is fictionally considered in the plot of the 2009 film ``Echelon Conspiracy''.} This is another way of phrasing the question posed in the title of the paper.

@@

This state of opacity affects a~large area of the phenomena of social life and raises concerns in some members of the global society. This opacity results in a~lack of information and—in a~sense—makes the entire area of reality epistemically inaccessible. However, this situation is not unique because similar situations appear in the case of the area of knowledge about the spiritual realm, in particular about God, and also in the microworld studied by quantum mechanics. Although God and the microworld are radically different from the world in which we live, we can sometimes find certain similarities, which we call analogies, between them. It is when we find ourselves in a~situation which is analogous but epistemically limited with respect to some reality that we often use a~research method called the argument from similarity. Its scheme may look like this:

@@

\item An object (of type) $X$~and an object (of type) $Y$~are similar (which is symbolically denoted as $X\approx Y$);
\item The similarity follows from $P$;
\item Theorem $T$~holds about object $X$~(symbolically: $T(X)$);
\item Therefore: theorem $T$~holds about $Y$~(symbolically: $T(Y)$).

@@

It is worth noting that, in general, we may be dealing with two areas of reality, one of which, i.e. a~source to which object $X$~belongs, is well known to us, while the other, i.e. a~target to which object $Y$~belongs, is not well known to us. The reasons for this may vary, and in our case they are opacity or lack of information. Thanks to the similarity we have previously found, we believe that we can cognitively \textit{invade} area $Y$. More precisely, this means that we can transfer our previously acquired knowledge about $X$~into ‘knowledge' about $Y$\footnote{These issues of similarity and argument are closely related to the theory of analogy, but we will not address them here.}.

\begin{table}[H]
\begin{small}
\begin{tabular}{|p{.45\textwidth}|p{.45\textwidth}|}%
\hline%
AREA $A$

\begin{enumerate}[wide, labelwidth=!, labelindent=0pt]
\item Object $X$: game $G(A)$;
\end{enumerate}
Similarity $P$:

\begin{enumerate}[wide, labelwidth=!, labelindent=0pt, start=2]
\item Player \textbf{I};
\item Player \textbf{II};
\item The game consists of moves and leads to a~result.
\item Winning: the play belongs to set $A$;
\item Theorem $T(X)$ holds: the game $G(A)$ will be won by player~\textbf{I}.
\end{enumerate}
%\vfill\null
&
AREA $B$

\begin{enumerate}[wide, labelwidth=!, labelindent=0pt]
\item Object $Y$: using of a~program;
\end{enumerate}
Similarity $P$:

\begin{enumerate}[wide, labelwidth=!, labelindent=0pt,start=2]
\item Algorithm (program);
\item A~human;
\item It consists of actions (elementary steps) and leads to an effect.
\item Winning: what the algorithm (machine) ``wants'' is accomplished;
\item Theorem $T(Y)$ holds: what the algorithm has planned will happen.
\end{enumerate}
%\vfill\null
\\\hline
\end{tabular}
\end{small}
\caption{A~summary of the points on which the analogy is based.}

@@

\begin{tabular}{|p{.25\textwidth}|p{.65\textwidth}|}
\hline
Player I: &
$x_{0} \quad x_{1} \quad x_{2} \qquad \ldots \qquad \ldots$\\\hline
Player II: &
$\ \quad y_{0} \quad y_{1} \quad y_{2} \qquad \quad\ldots \qquad \ldots$\\\hline
\end{tabular}
\end{small}
\caption{A~graphic representation of the game.}

@@

We use the moves of both players to create one infinite sequence of the form: $ z := \langle x_{0}, y_{0}, x_{1}, y_{1}, \ldots \rangle$, which we call \textit{a play} in \textit{a~game} $G(A)$. Definitions of players' strategies play a~key role:
\begingroup\abovedisplayskip=3pt \belowdisplayskip=3pt
\begin{definition}
Strategy $\sigma$ for Player \textbf{I} is the function:
$$\sigma: \{s\in\omega^{<\omega} : |s| \text{ is even}\} \rightarrow \omega.\qedhere$$
\end{definition}
\begin{definition}
Strategy $\tau$ for Player \textbf{II} is the function:
$$\tau: \{s\in\omega^{<\omega} : |s| \text{ is odd}\} \rightarrow \omega.\qedhere$$
\end{definition}
\begin{definition}
Let $\sigma$ be a~strategy for Player \textbf{I}, and the sequence $y = \langle y_{0}, y_{1}, y_{2}, \ldots \rangle$ be an infinite sequence of \textit{moves} of Player \textbf{II}, then:
$$\sigma*y = \langle x_{0}, y_{0}, x_{1}, y_{1}, x_{2}, y_{2}, \ldots \rangle,$$
where:\\
$x_{0} = \sigma({\langle}{\rangle});$\\
$x_{i+1} = \sigma(\langle x_{0}, y_{0}, x_{1}, y_{1}, \ldots, x_{i}, y_{i} \rangle).$
\end{definition}
\begin{definition}
Let $\tau$ be a~strategy for Player \textbf{II}, and the sequence $x = \langle x_{0}, x_{1}, x_{2}, \ldots \rangle$ be an infinite sequence of \textit{moves} of Player \textbf{I}, then:
$$x*\tau = \langle x_{0}, y_{0}, x_{1}, y_{1}, x_{2}, y_{2}, \ldots \rangle,$$
where:\\
$y_{0} = \tau(\langle x_{0} \rangle);$\\
$y_{i+1} = \tau(\langle x_{0}, y_{0}, x_{1}, y_{1}, \ldots, x_{i}, y_{i}, x_{i+1} \rangle).$
\end{definition}
\endgroup

@@

If $A \subset \omega^{\omega}$ is a~pay-off set, then:
\begin{itemize}
\item Strategy $\sigma$ is \textbf{a winning strategy} for Player \textbf{I} in Game $G(A)$ iff for all $y\in\omega^{\omega}$, we have: $\sigma*y \in A$.
\item Strategy $\tau$ is \textbf{a winning strategy} for Player \textbf{II} in Game $G(A)$ iff for all $x\in \omega^{\omega}$, we have: $x*\tau \not\in A$.
\end{itemize}
With the above definitions, we can formulate the axiom of determinacy (\textbf{AD}):
\begin{itemize}
\item For each set $A\subset \omega^{\omega}$, Game $G(A)$ is determined i.e. exactly one of the players has a~winning strategy for $G(A)$.\qedsymbol
\end{itemize}
This axiom contradicts the axiom of choice in the sense that the axiom of choice implies the existence of an undetermined infinite game.

@@

\begin{theorem}
Player I~has a~winning strategy in the Banach-Mazur game $G^{**}(A)$ iff $A$~is comeager
%\label{ref:RND2pCm13cCKk}(Soare, 2016, p.213).
\emph{\parencite[][p.213]{soare_turing_2016}}.%
\end{theorem}
%\textbf{Theorem 7.2}. Player I~has a~winning strategy in the Banach-Mazur game G**(A) iff A~is comeager
%%\label{ref:RND2pCm13cCKk}(Soare, 2016, p.213).
%\parencite[][p.213]{soare_turing_2016}.%

@@

The meager and comeager sets relate to Baire space $\Box^\Box$. Intuitively ``comeager sets are large. They form a~filter, are dense, uncountable, and are closed under countable intersections. Meager sets are small. They form an ideal, and countable sets are meager.''
%\label{ref:RNDIAkiz5FrEH}(Soare, 2016, p.212).
\parencite[][p.212]{soare_turing_2016}. %
 I~mention this because the matter may be of interest to philosophers.
\begin{proposition}
Let $A=GR_{1}$, then there is a~winning strategy for Player \textbf{II}.
\end{proposition}
%\textbf{Corollary 7.4}. Let A=GR\textsubscript{1}, then there is a~winning strategy for Player \textbf{II}.
\begin{proof}
Let $A=GR_{1}$, i.e. $A$~is the set of all unary generally recursive functions. The set $GR_{1}$ is countable. Therefore, by virtue of theorem \ref{olsz71}, there exists a~winning strategy for Player \textbf{II}.
\end{proof}
%Proof:
%
%Let A=GR\textsubscript{1}, i.e. A~is the set of all unary generally recursive functions. The set GR\textsubscript{1} is countable. Therefore, by virtue of theorem 7.1., there exists a~winning strategy for Player \textbf{II}.[F07F?]

>>>>

\begin{artengenv}{Adam Olszewski}
	{Will a~human always outsmart a~computer? An essay}
	{Will a~human always outsmart a~computer? An essay}
	{Will a~human always outsmart a~computer? An essay}
	{Pontifical University of John Paul II in Krakow}
	{The title question of the paper has its empirical origin in the form of an individual's existential experience arising from the personal use of a~computer, which we attempt to describe in the first section. The rest of the entire paper can be understood as a~philosophical essay answering the question posed. First the connection between the main problem of the article and its ``premonition'' by mankind, which was expressed in the form of ancient myths and legends, is briefly suggested. After shortly discussing the problems that early considerations of AI focused on, i.e. whether machines can think at all, we move on to reformulate our title question, about the possibility of outsmarting AI. This outsmarting will be understood by us in a~rather limited way as to prevent a~machine from completing its implemented task. To achieve this objective, after softly clarifying the basic terms, an analogy is built between the ``outsmarting'' of a~machine by a~human (the target domain) and the playing of a~mathematical game between two players (the base domain), where this outsmarting is assigned a~``winning strategy'' in the certain game. This mathematical model is formed by games similar to Banach-Mazur games. The strict theorems of such games are then proved and applied to the target of the analogy. We then draw conclusions and look for counter-examples to our findings. The answer to the title question posed is negative, and it is not clear how far it should be taken seriously.
	}
	{existential experience, myth, computer, machine, Banach-Mazur games, winning strategy.}

@@

\section{An existential ambient of the problem}
\lettrine[loversize=0.13,lines=2,lraise=-0.03,nindent=0em,findent=0.2pt]%
{T}{}he appearance of personal computers---which took place in the middle of 1970s in the world (USA) and a~decade later in Poland (due to the ``iron curtain'' which at that time separated this country from the rest of the world)---brought hope for providing man with a~useful tool with versatile applications and capable of performing certain very practical functions. Thus, in those days, from the perspective of an average user, a~computer standing on his desk was a~\textit{machine}, that is a~concrete, experiential and \textit{human-friendly} device, supporting his activities, for example as a~typewriter or a~memory bank for storing data. Since then, over nearly fifty years, such a~notion of a~useful machine has considerably evolved. This has happened thanks to the extremely dynamic development of information technology (which included increasing the computational capabilities of machines, their speed and memory and developing new and better programming techniques and languages) and the emergence of the Internet. It seems that the present concept of a~\textit{machine} has undergone a~substantial modification. This machine, which was a~friendly tool supporting man and operating in virtual (non-real) time and separated from the outside world, suddenly became a~machine operating in real time and additionally started drawing us deeper and deeper into a~kind of addiction and making us undertake certain activities, which---although performed through a~machine standing on the desk---have far-reaching consequences in real time and, more importantly, in reality. This new machine on our desk is, in fact, a~terminal of an extremely vast network (\textit{inter-net}), in which powerful algorithms create the elements of the \textit{machine proper}. This \textit{new machine}, which has taken on \textit{a~new meaning}, is also interactive in a~double sense. First, it is able to learn in a~certain sense strictly defined by computability theory, and second, it is known that it must be managed by or is in the hands of a~person or a~group of people because it is not an independent entity (or at least we do not know about it yet). We can call such a~machine \textit{artificial intelligence} (AI). We deliberately omit here a~wide array of issues that need to be clarified or systematically listed\footnote{For some clarifications see
%\label{ref:RNDasnA14J7h1}(Krzanowski and Polak, 2022).
\parencite[][]{krzanowski_meta-ontology_2022}.%
}; these shortcomings justify using the term \textit{essay} in the title of this work. Besides, this paper is my first attempt at analyzing this somewhat strange subject, so I~consider it to be the first of a~series of papers and an essayistic outline of my further work on the issue.

@@

\begin{itemize}
\item \textit{ante rem}---in the world of Platonic ideas---an algorithm;
\item \textit{post-ante rem}---as a~construction of the mind---a program;\footnote{\textit{Post-ante} is a~strange term which is to indicate that from one point of view a~program appears before the \textit{in re} phase and after the \textit{ante rem} phase, while from another point of view it appears after the \textit{in re} phase.}
\item \textit{in re}---physically present in the thing---implemented into a~machine (computer) in the form of the processes that control the behavior of the physical device.\footnote{Implementation is an extremely interesting concept in the philosophy of computer science. It allows us to transform something abstract into something physical, which requires thorough consideration. There is a~certain similarity here with the criticism of Platonism, where one of the objections against Platonism is the impossibility to move from the abstract to the physical.}
\end{itemize}
In other words, the three objects under consideration are the same object which manifests itself in three different forms or phases. We are not necessarily going to defend the above distinction at all costs, but we believe that it is easy to understand and will facilitate further reasoning although it will not be essential for the core of our considerations.

@@

\section{Myths as 'premonitions' of mankind}
The existential situation outlined above may---and sometimes does---cause anxiety in some members of the community of human users, as is amply demonstrated in literature. However, it is difficult to point to any facts to which we have access in the form of empirical verification, on the basis of which a~rigorous narrative could be constructed regarding the justification of existential anxiety. On the argumentative side, it is difficult to find such premises that would make it possible to justify a~conclusion regarding a~future threat from AI. Therefore, literary works devoted to this issue are based on speculations and some even treat it as science fiction, while others treat it as a~purely logical possibility (cf. article motto). We, however, do not downplay this premonition of humanity (expressed in literary terms), and in this section we will try to explain where this anxiety may come from. Our explanation refers to myths that appeared in human history a~long time ago. Mankind, as a~species, through its representatives, has created strange stories called \textit{myths}. I~call these stories strange because, being created at a~very early stage in the development of our species, they speak of problems that have been continuously accompanied it in its history. For example, we have the myth of the Sphinx, a~hybrid winged creature with the body of a~lion and the head of a~woman, who killed travelers heading to Thebes if they failed to solve a~riddle. A~similar pattern can be seen in the case of the dragon of Wawel Castle from the legend of Krakus and of many other creatures from various legends. The motif shared by many of these stories is an attempt to defeat or outwit an evil creature, which poses a~threat to the normal functioning of the community in which it appears. The origin of this creature is in some sense beyond the natural. Most often, in these stories an attempt to outwit it is a~rational act, such as solving the riddle of the Sphinx or giving a~dragon a~fake sheep filled with sulfur. Some believe---and these are not only so-called \textit{ordinary people} but also distinguished scientists---that in certain situations AI is \textit{per analogiam} appears as the embodiment of a~\textit{mythical creature}, which threatens the normal functioning of a~society, violating the freedom or privacy of all or some of its members. And, consequently, there is the question of outwitting the creature. Let us repeat that these myths, being a~kind of common heritage of mankind, are the real cause of humans' fears and anxieties.

\section{Turing's and Searle's tests---AI's first issues}
In 1936, the groundbreaking year in the creation of AI,\footnote{Here we are making a~mental shortcut, because strictly speaking it was crucial for AI to create a~computer as a~machine that practically realises the mathematical idea of computability.} two formal models of computability of effectively computable functions were published: one in the work by the American mathematician Alonzo Church (lambda calculus) and the other in the work by the American logician, born in Augustów (Poland) Emil Post (machine). A~year later the most famous work in the area was published: the one written by the English mathematician and logician, Alan Turing (machine). Very soon Turing's theoretical machines found implementation in the form of real machines, which were called \textit{computers}. This started a~discussion about the capabilities of such machines. Turing
%\label{ref:RNDhKwggb6xjl}(1950)
\parencite*[][]{turing_computing_1950} %
 proposed what is known today as \textit{the Turing test}, in which a~human evaluator judges whether his interlocutor in a~conversation held in a~natural language is a~human or a~machine (AI). Turing's intention in this test was to try to answer the question whether a~machine can think like a~human. He considered the theological argument (one of the arguments against machine thinking), according to which God created man as the only thinking being in the universe, and thinking was a~function of the human soul. Without going into the intricacies of the problem here, let us note that according to Catholic theology, thinking belongs to God and to His angels, thus it is not a~function of the body or brain as I~think Turing probably believed.\footnote{From a~certain point of view, calculation does not intrinsically belong to thinking, because, for example, according to Catholic theology, God, although he thinks, does not calculate, because calculation is the manifestation of a~certain kind of ignorance. Cf. for example Isaiah 55:8-9: ``For my thoughts are not your thoughts, neither are your ways my ways'', declares the Lord. ``As the heavens are higher than the earth, so are my [\ldots] thoughts [higher] than your thoughts.'' Some people believe (basing this belief on the Bible) that God's words in which He speaks of cause and effect justify attributing thinking to Him.} Turing wanted to convince his contemporaries that a~machine can think like a~human being in order to contradict underestimation of a~machine's capabilities in this regard, widespread at that time. The second test---called the Chinese Room argument---comes from John Searle 
%\label{ref:RND4igiqQ1wVO}(1980),
\parencite*[][]{searle_minds_1980}, %
 and was intended to demonstrate that no digital computer has a~kind of ``mind'' or ``consciousness'' even if it functionally bears a~far-reaching resemblance to human conversational behaviour. Searle's argument dampened the enthusiasm of optimistic proponents of AI and its unlimited possibilities. It can be said that the texts mentioned in this section share a~common goal: to take a~stand in the argument about machine thinking at a~time when, in popular understanding, a~machine's skills were not appreciated. From the point of view of this paper, both texts (Turing's and Searle's) have lost a~great deal of relevance since their publication, as so much has changed in this area. Summing up what has been said so far briefly and succinctly: today no one asks whether machines can think but rather what machines can do in terms of thinking and intelligence and where the upper limits of their capabilities lie.

\section{Preliminary formulation of the fundamental question}
Taking into account what has been outlined above, we can say that, from a~particular point of view, AI can appear as an element of reality---as an artefact---which poses a~threat to society on the way to the unrestricted realization of its development. Therefore, we should look for means by which we could ‘outsmart' this \textit{creature}. In different words, we can ask if a~human can stand up to AI that controls him.\footnote{An extreme case of this is fictionally considered in the plot of the 2009 film ``Echelon Conspiracy''.} This is another way of phrasing the question posed in the title of the paper.

@@

This state of opacity affects a~large area of the phenomena of social life and raises concerns in some members of the global society. This opacity results in a~lack of information and---in a~sense---makes the entire area of reality epistemically inaccessible. However, this situation is not unique because similar situations appear in the case of the area of knowledge about the spiritual realm, in particular about God, and also in the microworld studied by quantum mechanics. Although God and the microworld are radically different from the world in which we live, we can sometimes find certain similarities, which we call analogies, between them. It is when we find ourselves in a~situation which is analogous but epistemically limited with respect to some reality that we often use a~research method called the argument from similarity. Its scheme may look like this:

@@

\item An object (of type) X~and an object (of type) Y~are similar (which is symbolically denoted as X $\approx$ Y);
\item The similarity follows from P;
\item Theorem T~holds about object X~(symbolically: T(X));
\item Therefore: theorem T~holds about Y~(symbolically: T(Y)).

@@

It is worth noting that, in general, we may be dealing with two areas of reality, one of which, i.e. a~source to which object X~belongs, is well known to us, while the other, i.e. a~target to which object Y~belongs, is not well known to us. The reasons for this may vary, and in our case they are opacity or lack of information. Thanks to the similarity we have previously found, we believe that we can cognitively \textit{invade} area Y. More precisely, this means that we can transfer our previously acquired knowledge about X~into ‘knowledge' about Y\footnote{These issues of similarity and argument are closely related to the theory of analogy, but we will not address them here.}.

\begin{table}[H]
\begin{small}
\begin{tabular}{|p{.45\textwidth}|p{.45\textwidth}|}%
\hline%
AREA A

\begin{enumerate}[wide, labelwidth=!, labelindent=0pt]
\item Object X: game $G(A)$;
\end{enumerate}
Similarity P:

\begin{enumerate}[wide, labelwidth=!, labelindent=0pt, start=2]
\item Player \textbf{I};
\item Player \textbf{II};
\item The game consists of moves and leads to a~result.
\item Winning: the play belongs to set $A$;
\item Theorem T(X) holds: the game $G(A)$ will be won by Player~\textbf{I}.
\end{enumerate}
%\vfill\null
&
AREA B

\begin{enumerate}[wide, labelwidth=!, labelindent=0pt]
\item Object Y: using of a~program;
\end{enumerate}
Similarity P:

\begin{enumerate}[wide, labelwidth=!, labelindent=0pt,start=2]
\item Algorithm (program);
\item A~human;
\item It consists of actions (elementary steps) and leads to an effect.
\item Winning: what the algorithm (machine) ``wants'' is accomplished;
\item Theorem T(Y) holds: what the algorithm has planned will happen.
\end{enumerate}
%\vfill\null
\\\hline
\end{tabular}
\end{small}
\caption{A~summary of the points on which the analogy is based.}

@@

\begin{tabular}{|p{.25\textwidth}|p{.65\textwidth}|}
\hline
Player \textbf{I}: &
$x_{0} \quad x_{1} \quad x_{2} \qquad \ldots \qquad \ldots$\\\hline
Player \textbf{II}: &
$\ \quad y_{0} \quad y_{1} \quad y_{2} \qquad \quad\ldots \qquad \ldots$\\\hline
\end{tabular}
\end{small}
\caption{A~graphic representation of the game.}

@@

We use the moves of both players to create one infinite sequence of the form: $ z := \langle x_{0}, y_{0}, x_{1}, y_{1}, \ldots \rangle$, which we call \textit{a play} in \textit{a~game} $G(A)$. Definitions of players' strategies play a~key role:
\begingroup\abovedisplayskip=3pt \belowdisplayskip=3pt
\begin{definition}
Strategy $\sigma$ for Player \textbf{I} is the function:
$$\sigma: \{s\in\omega^{<\omega} : |s| \text{ is even}\} \rightarrow \omega.\qedhere$$
\end{definition}
\begin{definition}
Strategy $\tau$ for Player \textbf{II} is the function:
$$\tau: \{s\in\omega^{<\omega} : |s| \text{ is odd}\} \rightarrow \omega.\qedhere$$
\end{definition}
\begin{definition}
Let $\sigma$ be a~strategy for Player \textbf{I}, and the sequence $y = \langle y_{0}, y_{1}, y_{2}, \ldots \rangle$ be an infinite sequence of \textit{moves} of Player \textbf{II}, then:
$$\sigma^*y = \langle x_{0}, y_{0}, x_{1}, y_{1}, x_{2}, y_{2}, \ldots \rangle,$$
where:\\
$x_{0} = \sigma({\langle}{\rangle});$\\
$x_{i+1} = \sigma(\langle x_{0}, y_{0}, x_{1}, y_{1}, \ldots, x_{i}, y_{i} \rangle).$
\end{definition}
\begin{definition}
Let $\tau$ be a~strategy for Player \textbf{II}, and the sequence $x = \langle x_{0}, x_{1}, x_{2}, \ldots \rangle$ be an infinite sequence of \textit{moves} of Player \textbf{I}, then:
$$x^*\tau = \langle x_{0}, y_{0}, x_{1}, y_{1}, x_{2}, y_{2}, \ldots \rangle,$$
where:\\
$y_{0} = \tau(\langle x_{0} \rangle);$\\
$y_{i+1} = \tau(\langle x_{0}, y_{0}, x_{1}, y_{1}, \ldots, x_{i}, y_{i}, x_{i+1} \rangle).$
\end{definition}
\endgroup

@@

If $A \subset \omega^{\omega}$ is a~pay-off set, then:
\begin{itemize}
\item Strategy $\sigma$ is \textbf{a winning strategy} for Player \textbf{I} in Game $G(A)$ iff for all $y\in\omega^{\omega}$, we have: $\sigma^*y \in A$.
\item Strategy $\tau$ is \textbf{a winning strategy} for Player \textbf{II} in Game $G(A)$ iff for all $x\in \omega^{\omega}$, we have: $x^*\tau \not\in A$.
\end{itemize}
With the above definitions, we can formulate the axiom of determinacy (\textbf{AD}):
\begin{itemize}
\item[(\textbf{AD})] For each set $A\subset \omega^{\omega}$, Game $G(A)$ is determined i.e. exactly one of the players has a~winning strategy for $G(A)$.\qedsymbol
\end{itemize}
This axiom contradicts the axiom of choice in the sense that the axiom of choice implies the existence of an undetermined infinite game.

@@

\begin{theorem}
Player \textbf{I}~has a~winning strategy in the Banach-Mazur game $G^{**}(A)$ iff $A$~is comeager
%\label{ref:RND2pCm13cCKk}(Soare, 2016, p.213).
\emph{\parencite[][p.213]{soare_turing_2016}}.%
\end{theorem}
%\textbf{Theorem 7.2}. Player I~has a~winning strategy in the Banach-Mazur game G**(A) iff A~is comeager
%%\label{ref:RND2pCm13cCKk}(Soare, 2016, p.213).
%\parencite[][p.213]{soare_turing_2016}.%

@@

The meager and comeager sets relate to Baire space $\Box^\Box$. Intuitively ``comeager sets are large. They form a~filter, are dense, uncountable, and are closed under countable intersections. Meager sets are small. They form an ideal, and countable sets are meager.''
%\label{ref:RNDIAkiz5FrEH}(Soare, 2016, p.212).
\parencite[][p.212]{soare_turing_2016}. %
 I~mention this because the matter may be of interest to philosophers.
\begin{proposition}
Let $A=GR_{1}$, then there is a~winning strategy for Player \textbf{II}.
\end{proposition}
%\textbf{Corollary 7.4}. Let A=GR\textsubscript{1}, then there is a~winning strategy for Player \textbf{II}.
\begin{proof}
Let $A=GR_{1}$, i.e. $A$~is the set of all unary generally recursive functions. The set $GR_{1}$ is countable. Therefore, by virtue of theorem~\ref{olsz71}, there exists a~winning strategy for Player \textbf{II}.
\end{proof}
%Proof:
%
%Let A=GR\textsubscript{1}, i.e. A~is the set of all unary generally recursive functions. The set GR\textsubscript{1} is countable. Therefore, by virtue of theorem 7.1., there exists a~winning strategy for Player \textbf{II}.[F07F?]
